Kalman tracking and estimation of continuous gravitational waves with a pulsar timing array

### 2023-08-30 docs/manuscript/paper.tex
bayes for multiple noise realisations

\begin{figure*}
	\includegraphics[width=\textwidth, height =\textwidth ]{images/representative_example_v2_GW}
	\caption{Posterior distribution for the GW source parameters $\boldsymbol{\theta}_{\rm gw}$ for the representative system described in Table \ref{tab:parameters_and_priors}, for a single realisation of the system noise. The vertical orange lines indicate the true injected values. The contours in the 2D histograms denote the (0.5, 1, 1.5, 2)-$\sigma$ levels. We are able to accurately estimate each parameter of interest. We pot the scaled variables $\Omega_{\rm nHz} = \Omega \times 10^{9}$ Hz and $h_{0, \times 10^{12}} = h_0 \times 10^{12}$.}
	\label{fig:corner_plot_1}
\end{figure*}

@@

\subsubsection{Multiple noise realisations} \label{sec:multiple_noise}
%\begin{figure*}
%	\includegraphics[width=\textwidth, height =\textwidth]{images/stacked_GW_plot_1000}
%	\caption{As Figure \ref{fig:corner_plot_1} but for 9 realisations of the noise processes. The inferred posterior distributions show strong agreement between the different noise realisations for the majority of static parameters. The exceptions are $\iota$ and $h$ which exhibit a larger degree of variance due to the weak identifiability between these two parameters c.f. Figure \ref{fig:likelihood_surface}.} 
%	\label{fig:corner_plot_2}
%\end{figure*}

\begin{figure*}
	%\centering % Not needed
	\begin{subfigure}[b]{0.49\textwidth}
		\includegraphics[width=\textwidth]{images/stacked_GW_plot_1000}
		\caption{Kalman filter using correct estimate of the static parameters, $\hat{\boldsymbol{\theta}} = \boldsymbol{\theta}$ }
		\label{fig:6MB_BFS}
	\end{subfigure}  
	\hfill
	\begin{subfigure}[b]{0.49\textwidth}
		\includegraphics[width=\textwidth]{images/stacked_GW_plot_1000}
		\caption{Kalman filter using incorrect estimate of the static parameters, $\hat{\boldsymbol{\theta}} \neq \boldsymbol{\theta}$}
		\label{fig:25MB_bfs}
	\end{subfigure}
	\caption{As Figure \ref{fig:corner_plot_1} but for 9 realisations of the noise processes. The inferred posterior distributions show strong agreement between the different noise realisations for the majority of static parameters. The exceptions are $\iota$ and $h$ which exhibit a larger degree of variance due to the weak identifiability between these two parameters c.f. Figure \ref{fig:likelihood_surface}.}
	\label{fig:four figures2}
\end{figure*}

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{images/likelihood_surface}
	\caption{Likelihood $\log \mathcal{L}$ (Equation \ref{eq:likelihood}) surface across the $\iota - h_0$ parameter space for a single noise realisation. The red point labels the location of the likelihood maxima. The likelihood surface has been normalised with respect to the absolute value of the maxima. A strong likelihood ridge is evident in the parameter space where the Kalman filter produces similar likelihood values, making it difficult for sampling algorithms to converge to a consistent value.}
	\label{fig:likelihood_surface}
\end{figure}
\begin{figure*}
	\centering
	\includegraphics[width=\textwidth]{images/compare_iotas_lines2}
	\caption{One-dimensional marginalized posteriors for $\iota$ over 9 realisations of the noise where $h$ is (a) an unknown free parameter (left panel) and (b) fixed at its true value $= 10^{-12}$ (right panel). The thick vertical orange line labels the true injected value. The dashed vertical lines denote the median of the corresponding coloured posterior. Fixing $h$ reduces the variance between different noise realisations but results in a clear bias between the true and inferred values of $\iota$. The left hand panel is equivalent to the marginalized posteriors for $\iota$ shown in Figure \ref{fig:corner_plot_2} (4th row, 4th column), but with the $(0.16,0.84)$ quantiles replaced with the $0.5$ quantile.}
	\label{fig:iota}
\end{figure*}
The above discussion was for a single realisation of the noise processes. It is important to confirm that our method is robust against different realisations of the noise and that the specific realisation of the noisy data used in Sections \ref{sec:parameter_estim} and \ref{sec:detection} is not particularly advantageous or convenient for our method. In this Section we confirm that we are able to infer comparable results for different noise realisations. To this end we take our representative example from Table \ref{tab:parameters_and_priors} and generate 1000 realisations of the process noise and measurement noise. For each realisation of the data we can then independently estimate the static parameters, $\boldsymbol{\theta}$. In Figure \ref{fig:corner_plot_2} we show the results for the  estimated values of $\boldsymbol{\theta}_{\rm gw}$ for nine realisations of the noise. This figure is analogous to the single noise realisation of Figure \ref{fig:corner_plot_1}. We plot only 9 realisations of the noise rather than the full set of 1000 realisations so as to not overcrowd the figure. As in Section \ref{sec:parameter_estim} the estimates of $\boldsymbol{\theta}_{\rm psr}$ are recovered unambiguously across the 1000 noise realisations and for brevity we do not show the results here. \newline

We now consider the complete set of $10^3$ noise realisations, rather than the subset of 9 realisations in the preceding discussion. For each realisation of the noise we infer a marginalised 1-D posterior for each parameter. For this injected strain ($h_0 = 10^{-12}$) the signal-to-noise ratio (SNR) is large (c.f. Figure \ref{fig:bayes}) and consequently the posteriors are typically narrow compared to the prior space and well described by their median value. To compare each of the posteriors across different noise realisations,

We saw previously strong biases in $\iota$ and $h_0$ when considering just 9 noise realisations. These biases persist with a similar magnitude when we consider the full set of $10^3$ noise realisations. Moreover, smaller biases are also observed in $\psi$ ($\sim 0.1$ radians) and $\alpha$ ($\sim 0.04$ radians) as well as smaller still biases in $\delta$ and $\Phi_0$. These biases are a result of dropping the pulsar terms from the measurement equation and will be discussed in Section \ref{sec:bias}
\subsection{Exploring a broader parameter space} \label{sec:parameter_space}
\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{images/pp_plot_new}
	\caption{PP plot for 200 simulated GW sources for a subset of parameters of $\boldsymbol{\theta}_{\rm gw}$, randomly drawn from the prior distributions of Table \ref{tab:parameters_and_priors}. Each coloured line corresponds to a different parameter of $\boldsymbol{\theta}_{\rm gw}$. The parameters $h_0$ and $\iota$ are fixed at $5 \times 10^{-15}$ and 1.0 rad in order to maintain an approximately constant SNR. The grey shaded contours labels the $1,2,3$-$\sigma$ confidence intervals. Well estimated posteriors should fall along the diagonal $y=x$. Many parameters show evidence of being over-constrained due to a modelling bias. }
	\label{fig:parameter_space}
\end{figure}
We have so far focussed on a single representative system, summarised in Table \ref{tab:parameters_and_priors}. It is also important to test the method in different regions of the parameter space. To this end we consider 200 injections for different sets of parameter values where we fix $h_0 = 5 \times 10^{-15}$, $\iota =1.0$ and draw the remaining five parameters of $\boldsymbol{\theta}_{\rm gw}$ from the prior distributions described in Table \ref{tab:parameters_and_priors}. We fix $h_0$ and $\iota$ in order to maintain an approximately constant SNR across the parameter space. For each simulated injection we can then attempt to recover the posterior distributions for each parameter. To summarise the results across the parameter space we use a parameter-parameter (PP) plot \citep{doi:10.1198/106186006X136976}. A PP plot describes the fraction of the total number of injected parameters which are included within a given credible interval of the estimated posterior with respect to the credible interval itself. In the ideal case the PP plot should be a diagonal line, indicating that $x$-percentage of the simulated injections fall within the $x$-percentage credible interval. The results are shown in Figure \ref{fig:parameter_space}. The shaded grey contours enclose the $1\sigma$, $2\sigma$, and $3\sigma$ significance levels, given 200 injections."
We can see that only $\Omega$ falls within the $3\sigma$ shaded region. The other parameters deviate from the $y=x$ diagonal. The effect is most pronounced for $\alpha$ and $\delta$ with more modest deviations for $\psi$ and $\Phi_0$. The shape of the graph indicates that the posteriors for these parameters are over-constrained; there are fewer injections contained within high value credible intervals than would be expected statistically, and there are more injections contained within the low value credible intervals. This is an result of the bias that we observed in Section \ref{sec:multiple_noise}; the posteriors inferred for these parameters are generally highly confident, i.e. narrow, but are systematically biased away from the true injection values. This manifests in the PP plot as posteriors which are overly precise (narrow) to contain the injected value within the appropriate credible interval. An in-depth discussion on the origin of this bias is given in Section \ref{sec:bias}.

@@

\section{Wasserstein distance}
In order to compare the control and competitor ensemble probability distributions we adopt as our key metric the Wasserstein distance (WD). The WD intuitively defines a distance between two distributions as the optimal cost, with respect to some cost function $c(x,y)$,  of transporting a probability mass from position $x$ to position $y$. The $p$-Wasserstein distance between two distributions $\mu, \nu$ is
\begin{eqnarray}
	W_p(\mu,\nu)= \left( \inf_{\gamma \in \Gamma(\mu, \nu)}  \int c(x,y)^p d \gamma (x,y)\right)^{1/p}
\end{eqnarray}
where $\gamma(x,y)$ is the transport plan, $\Gamma(\mu, \nu)$ is the set of all couplings of $\mu$ and $\nu$ and $c(x,y)$ is the cost function. It is well-known \cite{Dudley} that the Wasserstein distance suffers from the curse of dimensionality for dimensions $d \geq 3$:
\begin{eqnarray}
	\boldsymbol{E} [W_1(\mu_n,\mu)] \asymp n^{-\frac{1}{d}}
\end{eqnarray}
To sidestep this issue, we marginalize onto the distributions spanned by the individual gridpoints and take the WD between these 1-dimensional distributions. Such an approach is also adopted in \cite{Paxton2021,Vissio2020}. We also take $p=1$ and $c(x,y) = |x-y|$ such that the WD values quoted in this work are given by
\begin{eqnarray}
	W_1(\mu,\nu)=  \inf_{\gamma \in \Gamma(\mu, \nu)} \int |x-y| d \gamma (x,y)
\end{eqnarray}
From the perspective of climate modelling, the WD has a nice physical interpretation given by the Monge-Kantorovich duality:
\begin{eqnarray}
	| \boldsymbol{E}(X_{\mu} )-\boldsymbol{E}(Y_{\nu} ) | \leq W_1(\mu, \nu) \label{eq:WDdefn}
\end{eqnarray}
i.e. the difference in the expected value of some random variable between two distributions is no bigger than the WD value itself. More explicitly, if the WD for surface temperature between two distributions is 1K, then this tells us that the difference in the expected temperature between the distributions is no greater than 1K. For a full review of the Wasserstein distance, including favourable properties and comparison with other metrics we refer the reader to the Appendix of Pax21.

>>>>

\begin{figure*}
	\includegraphics[width=\textwidth, height =\textwidth ]{images/small_h_posterior_10}
	\caption{Posterior distribution for the GW source parameters $\boldsymbol{\theta}_{\rm gw}$ for the representative system described in Table \ref{tab:parameters_and_priors}, for a single realisation of the system noise. The vertical orange lines indicate the true injected values. The contours in the 2D histograms denote the (0.5, 1, 1.5, 2)-$\sigma$ levels. We are able to accurately estimate each parameter of interest. We pot the scaled variables $\Omega_{\rm nHz} = \Omega \times 10^{9}$ Hz and $h_{0, \times 10^{12}} = h_0 \times 10^{12}$.}
	\label{fig:corner_plot_1}
\end{figure*}

%\begin{figure*}
%	\includegraphics[width=\textwidth, height =\textwidth ]{images/representative_example_v2_GW}
%	\caption{Posterior distribution for the GW source parameters $\boldsymbol{\theta}_{\rm gw}$ for the representative system described in Table \ref{tab:parameters_and_priors}, for a single realisation of the system noise. The vertical orange lines indicate the true injected values. The contours in the 2D histograms denote the (0.5, 1, 1.5, 2)-$\sigma$ levels. We are able to accurately estimate each parameter of interest. We pot the scaled variables $\Omega_{\rm nHz} = \Omega \times 10^{9}$ Hz and $h_{0, \times 10^{12}} = h_0 \times 10^{12}$.}
%	\label{fig:corner_plot_1}
%\end{figure*}

\subsubsection{Multiple noise realisations} \label{sec:multiple_noise}
\begin{figure*}
	\includegraphics[width=\textwidth, height =\textwidth]{images/stacked_GW_plot_1000}
	\caption{As Figure \ref{fig:corner_plot_1} but for 9 realisations of the noise processes. The inferred posterior distributions show strong agreement between the different noise realisations for the majority of static parameters. The exceptions are $\iota$ and $h$ which exhibit a larger degree of variance due to the weak identifiability between these two parameters c.f. Figure \ref{fig:likelihood_surface}.} 
	\label{fig:corner_plot_2}
\end{figure*}

%\begin{figure*}
%	%\centering % Not needed
%	\begin{subfigure}[b]{0.49\textwidth}
%		\includegraphics[width=\textwidth]{images/stacked_GW_plot_1000}
%		\caption{Kalman filter using correct estimate of the static parameters, $\hat{\boldsymbol{\theta}} = \boldsymbol{\theta}$ }
%		\label{fig:6MB_BFS}
%	\end{subfigure}  
%	\hfill
%	\begin{subfigure}[b]{0.49\textwidth}
%		\includegraphics[width=\textwidth]{images/stacked_GW_plot_1000}
%		\caption{Kalman filter using incorrect estimate of the static parameters, $\hat{\boldsymbol{\theta}} \neq \boldsymbol{\theta}$}
%		\label{fig:25MB_bfs}
%	\end{subfigure}
%	\caption{As Figure \ref{fig:corner_plot_1} but for 9 realisations of the noise processes. The inferred posterior distributions show strong agreement between the different noise realisations for the majority of static parameters. The exceptions are $\iota$ and $h$ which exhibit a larger degree of variance due to the weak identifiability between these two parameters c.f. Figure \ref{fig:likelihood_surface}.}
%	\label{fig:four figures2}
%\end{figure*}

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{images/likelihood_surface}
	\caption{Likelihood $\log \mathcal{L}$ (Equation \ref{eq:likelihood}) surface across the $\iota - h_0$ parameter space for a single noise realisation. The red point labels the location of the likelihood maxima. The likelihood surface has been normalised with respect to the absolute value of the maxima. A strong likelihood ridge is evident in the parameter space where the Kalman filter produces similar likelihood values, making it difficult for sampling algorithms to converge to a consistent value.}
	\label{fig:likelihood_surface}
\end{figure}
%\begin{figure*}
%	\centering
%	\includegraphics[width=\textwidth]{images/compare_iotas_lines2}
%	\caption{One-dimensional marginalized posteriors for $\iota$ over 9 realisations of the noise where $h$ is (a) an unknown free parameter (left panel) and (b) fixed at its true value $= 10^{-12}$ (right panel). The thick vertical orange line labels the true injected value. The dashed vertical lines denote the median of the corresponding coloured posterior. Fixing $h$ reduces the variance between different noise realisations but results in a clear bias between the true and inferred values of $\iota$. The left hand panel is equivalent to the marginalized posteriors for $\iota$ shown in Figure \ref{fig:corner_plot_2} (4th row, 4th column), but with the $(0.16,0.84)$ quantiles replaced with the $0.5$ quantile.}
%	\label{fig:iota}
%\end{figure*}
The above discussion was for a single realisation of the noise processes. It is important to confirm that our method is robust against different realisations of the noise and that the specific realisation of the noisy data used in Sections \ref{sec:parameter_estim} and \ref{sec:detection} is not particularly advantageous or convenient for our method. In this Section we confirm that we are able to infer comparable results for different noise realisations. To this end we take our representative example from Table \ref{tab:parameters_and_priors} and generate 1000 realisations of the process noise and measurement noise. For each realisation of the data we can then independently estimate the static parameters, $\boldsymbol{\theta}$. In Figure \ref{fig:corner_plot_2} we show the results for the  estimated values of $\boldsymbol{\theta}_{\rm gw}$ for nine realisations of the noise. This figure is analogous to the single noise realisation of Figure \ref{fig:corner_plot_1}. We plot only 9 realisations of the noise rather than the full set of 1000 realisations so as to not overcrowd the figure. As in Section \ref{sec:parameter_estim} the estimates of $\boldsymbol{\theta}_{\rm psr}$ are recovered unambiguously across the 1000 noise realisations and for brevity we do not show the results here. \newline

We now consider the complete set of $10^3$ noise realisations, rather than the subset of ten realisations in the preceding discussion. For each realisation of the noise we infer a marginalised one-dimensional posterior for each parameter. We want to compare these posteriors for different noise realisations and evaluate whether our method can consistently and reliably converge to comparable posteriors.

To this end we use the the Wasserstein distance \citep[WD,][]{Wasserstein,Villani2009} from optimal transport theory which defines an intuitive of similarity between probability distributions.

A full review of the Wasserstein distance which may be unfamiliar to some readers, is presented in Appendix \ref{sec:wasserstein}

In order to quantitatively define `similar'  we use . The WD is a metric which quantifies the distance between two probability distributions 
as the optimal cost, with respect to some cost function $c(x,y)$,  of transporting a probability mass from position $x$ to position $y$. The $p$-Wasserstein distance between two distributions $\mu, \nu$ is
\begin{eqnarray}
	W_p(\mu,\nu)= \left( \inf_{\gamma \in \Gamma(\mu, \nu)}  \int c(x,y)^p d \gamma (x,y)\right)^{1/p}
\end{eqnarray}
where $\gamma(x,y)$ is the transport plan, $\Gamma(\mu, \nu)$ is the set of all couplings of $\mu$ and $\nu$ and $c(x,y)$ is the cost function.

In order to compare the control and competitor ensemble probability distributions we adopt as our key metric the Wasserstein distance (WD). The WD intuitively defines a distance between s

We also take $p=1$ and $c(x,y) = |x-y|$ such that the WD values quoted in this work are given by
\begin{eqnarray}
	W_1(\mu,\nu)=  \inf_{\gamma \in \Gamma(\mu, \nu)} \int |x-y| d \gamma (x,y)
\end{eqnarray}
From the perspective of climate modelling, the WD has a nice physical interpretation given by the Monge-Kantorovich duality:
\begin{eqnarray}
	| \boldsymbol{E}(X_{\mu} )-\boldsymbol{E}(Y_{\nu} ) | \leq W_1(\mu, \nu) \label{eq:WDdefn}
\end{eqnarray}
i.e. the difference in the expected value of some random variable between two distributions is no bigger than the WD value itself. More explicitly, if the WD for surface temperature between two distributions is 1K, then this tells us that the difference in the expected temperature between the distributions is no greater than 1K. For a full review of the Wasserstein distance, including favourable properties and comparison with other metrics we refer the reader to the Appendix of Pax21.

For this injected strain ($h_0 = 10^{-12}$) the signal-to-noise ratio (SNR) is large (c.f. Figure \ref{fig:bayes}) and consequently the posteriors are typically narrow compared to the prior space and well described by their median value. To compare each of the posteriors across different noise realisations,

We saw previously strong biases in $\iota$ and $h_0$ when considering just 9 noise realisations. These biases persist with a similar magnitude when we consider the full set of $10^3$ noise realisations. Moreover, smaller biases are also observed in $\psi$ ($\sim 0.1$ radians) and $\alpha$ ($\sim 0.04$ radians) as well as smaller still biases in $\delta$ and $\Phi_0$. These biases are a result of dropping the pulsar terms from the measurement equation and will be discussed in Section \ref{sec:bias}

@@

\subsection{Exploring a broader parameter space} \label{sec:parameter_space}
\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{images/pp_plot_new}
	\caption{PP plot for 200 simulated GW sources for a subset of parameters of $\boldsymbol{\theta}_{\rm gw}$, randomly drawn from the prior distributions of Table \ref{tab:parameters_and_priors}. Each coloured line corresponds to a different parameter of $\boldsymbol{\theta}_{\rm gw}$. The parameters $h_0$ and $\iota$ are fixed at $5 \times 10^{-15}$ and 1.0 rad in order to maintain an approximately constant SNR. The grey shaded contours labels the $1,2,3$-$\sigma$ confidence intervals. Well estimated posteriors should fall along the diagonal $y=x$. Many parameters show evidence of being over-constrained due to a modelling bias. }
	\label{fig:parameter_space}
\end{figure}
We have so far focussed on a single representative system, summarised in Table \ref{tab:parameters_and_priors}. It is also important to test the method in different regions of the parameter space. To this end we consider 200 injections for different sets of parameter values where we fix $h_0 = 5 \times 10^{-15}$, $\iota =1.0$ and draw the remaining five parameters of $\boldsymbol{\theta}_{\rm gw}$ from the prior distributions described in Table \ref{tab:parameters_and_priors}. We fix $h_0$ and $\iota$ in order to maintain an approximately constant SNR across the parameter space. For each simulated injection we can then attempt to recover the posterior distributions for each parameter. To summarise the results across the parameter space we use a parameter-parameter (PP) plot \citep{doi:10.1198/106186006X136976}. A PP plot describes the fraction of the total number of injected parameters which are included within a given credible interval of the estimated posterior with respect to the credible interval itself. In the ideal case the PP plot should be a diagonal line, indicating that $x$-percentage of the simulated injections fall within the $x$-percentage credible interval. The results are shown in Figure \ref{fig:parameter_space}. The shaded grey contours enclose the $1\sigma$, $2\sigma$, and $3\sigma$ significance levels, given 200 injections."
We can see that only $\Omega$ falls within the $3\sigma$ shaded region. The other parameters deviate from the $y=x$ diagonal. The effect is most pronounced for $\alpha$ and $\delta$ with more modest deviations for $\psi$ and $\Phi_0$. The shape of the graph indicates that the posteriors for these parameters are over-constrained; there are fewer injections contained within high value credible intervals than would be expected statistically, and there are more injections contained within the low value credible intervals. This is an result of the bias that we observed in Section \ref{sec:multiple_noise}; the posteriors inferred for these parameters are generally highly confident, i.e. narrow, but are systematically biased away from the true injection values. This manifests in the PP plot as posteriors which are overly precise (narrow) to contain the injected value within the appropriate credible interval. An in-depth discussion on the origin of this bias is given in Section \ref{sec:bias}.

@@

\section{Wasserstein distance}\label{sec:wasserstein}

It is well-known \cite{Dudley} that the Wasserstein distance suffers from the curse of dimensionality for dimensions $d \geq 3$:
\begin{eqnarray}
	\boldsymbol{E} [W_1(\mu_n,\mu)] \asymp n^{-\frac{1}{d}}
\end{eqnarray}
To sidestep this issue, we marginalize onto the distributions spanned by the individual gridpoints and take the WD between these 1-dimensional distributions. Such an approach is also adopted in \cite{Paxton2021,Vissio2020}.

In order to compare the control and competitor ensemble probability distributions we adopt as our key metric the Wasserstein distance (WD). The WD intuitively defines a distance between two distributions as the optimal cost, with respect to some cost function $c(x,y)$,  of transporting a probability mass from position $x$ to position $y$. The $p$-Wasserstein distance between two distributions $\mu, \nu$ is
\begin{eqnarray}
	W_p(\mu,\nu)= \left( \inf_{\gamma \in \Gamma(\mu, \nu)}  \int c(x,y)^p d \gamma (x,y)\right)^{1/p}
\end{eqnarray}
where $\gamma(x,y)$ is the transport plan, $\Gamma(\mu, \nu)$ is the set of all couplings of $\mu$ and $\nu$ and $c(x,y)$ is the cost function. It is well-known \cite{Dudley} that the Wasserstein distance suffers from the curse of dimensionality for dimensions $d \geq 3$:
\begin{eqnarray}
	\boldsymbol{E} [W_1(\mu_n,\mu)] \asymp n^{-\frac{1}{d}}
\end{eqnarray}
To sidestep this issue, we marginalize onto the distributions spanned by the individual gridpoints and take the WD between these 1-dimensional distributions. Such an approach is also adopted in \cite{Paxton2021,Vissio2020}. We also take $p=1$ and $c(x,y) = |x-y|$ such that the WD values quoted in this work are given by
\begin{eqnarray}
	W_1(\mu,\nu)=  \inf_{\gamma \in \Gamma(\mu, \nu)} \int |x-y| d \gamma (x,y)
\end{eqnarray}
From the perspective of climate modelling, the WD has a nice physical interpretation given by the Monge-Kantorovich duality:
\begin{eqnarray}
	| \boldsymbol{E}(X_{\mu} )-\boldsymbol{E}(Y_{\nu} ) | \leq W_1(\mu, \nu) \label{eq:WDdefn}
\end{eqnarray}
i.e. the difference in the expected value of some random variable between two distributions is no bigger than the WD value itself. More explicitly, if the WD for surface temperature between two distributions is 1K, then this tells us that the difference in the expected temperature between the distributions is no greater than 1K. For a full review of the Wasserstein distance, including favourable properties and comparison with other metrics we refer the reader to the Appendix of Pax21.
